---
title: "Integrated Transcriptomic Analysis of Dilated and Ischemic Cardiomyopathy"
author: "Anas Jamshed"
date: today

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo

  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo

execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false

library(knitr)
library(ggplot2)
library(dplyr)


```

# Abstract

Cardiomyopathy is a major cause of heart failure and cardiovascular mortality worldwide. This study integrates six independent Gene Expression Omnibus (GEO) microarray datasets to investigate transcriptomic alterations associated with ischemic and non-ischemic cardiomyopathy.

The workflow involved GEO data acquisition, probe annotation harmonization, Entrez Gene ID mapping, multi-study integration, batch effect correction using ComBat, differential expression analysis using Limma, and downstream visualization using PCA, heatmaps, correlation matrices, and volcano plots.

# Introduction

Individual microarray studies often suffer from limited sample sizes and platform-specific biases. Integrating multiple independent datasets improves statistical power and biological robustness.

The following GEO datasets were analyzed:

| GEO Series | Platform |
|------------|----------|
| GSE1145    | GPL570   |
| GSE1869    | GPL96    |
| GSE3585    | GPL96    |
| GSE3586    | GPL3050  |
| GSE9128    | GPL96    |
| GSE42955   | GPL96    |

# Study Design

The analysis compared:

-   Control samples
-   Non-Ischemic Cardiomyopathy (DCM)
-   Ischemic Cardiomyopathy (ICM)

# Target Genes

The study focused on several cardiomyopathy-associated genes.

```{r}
target_genes <- c(
"ACTN2","CRYAB","BMP10",
"CSRP3","DES","FHOD3",
"FLNC","LDB3","MYZAP",
"MYPN","MYOZ2","NEXN",
"PDLIM3","PDLIM5",
"TCAP","TTN"
)

kable(
data.frame(Target_Gene = target_genes),
caption = "Cardiomyopathy Candidate Genes"
)
```

# Materials and Methods

## Data Acquisition

Public microarray datasets were downloaded from NCBI GEO using the GEOquery package.

## Gene Annotation

Probe identifiers were converted into Entrez Gene IDs using AnnotationDbi and org.Hs.eg.db.

## Dataset Integration

Expression matrices were aggregated at gene level and merged using common Entrez Gene IDs.

## Metadata Harmonization

Clinical labels from all studies were standardized into:

-   Control Group
-   Non-Ischemic CM
-   Ischemic CM

## Batch Effect Correction

To minimize study-specific variation, ComBat from the sva package was applied.

## Differential Expression Analysis

Differential expression analysis was performed using:

-   Limma
-   Voom normalization
-   Empirical Bayes moderation

## Visualization

The following visualizations were generated:

-   PCA
-   Heatmaps
-   Correlation Heatmaps
-   Volcano Plots
-   Enhanced Volcano Plots
-   Boxplots

# Differential Expression Results

## Non-Ischemic Cardiomyopathy vs Control

```{r}
dc <- read.csv(
"Differentially_expressed_genes_control_vs_Dilated_cardiomypaty.csv"
)

kable(
head(dc[,c("GENE_SYMBOL","logFC","adj.P.Val")],20),
caption = "Top Differentially Expressed Genes"
)
```

## Ischemic Cardiomyopathy vs Control

```{r}
icm <- read.csv(
"Differentially_expressed_genes_control_vs_ischemic_cardiomypaty.csv"
)

kable(
head(icm[,c("GENE_SYMBOL","logFC","adj.P.Val")],20),
caption = "Top Differentially Expressed Genes"
)
```

# Principal Component Analysis

Principal Component Analysis was performed after batch correction.

## Non-Ischemic Cardiomyopathy

```{r}
knitr::include_graphics(
"PCA_control_vs_Dilated_Cardiomyopaty.png"
)
```

## Ischemic Cardiomyopathy

```{r}
knitr::include_graphics(
"PCA_control_vs_Ischemic_Cardiomyopaty.png"
)
```

## Integrated Dataset

```{r}
knitr::include_graphics(
"PCA_All_data.png"
)
```

# Heatmap Analysis

## Control vs Non-Ischemic Cardiomyopathy

```{r}
knitr::include_graphics(
"control_vs_Dilated_heatmap.png"
)
```

## Control vs Ischemic Cardiomyopathy

```{r}
knitr::include_graphics(
"control_vs_ischemic_heatmap.png"
)
```

# Correlation Analysis

```{r}
knitr::include_graphics(
"corrheatmap.png"
)
```

# Volcano Plot Analysis

## Non-Ischemic Cardiomyopathy

```{r}
knitr::include_graphics(
"DC_volcano.png"
)
```

## Ischemic Cardiomyopathy

```{r}
knitr::include_graphics(
"Ischemic_volcano.png"
)
```

# Enhanced Volcano Plots

## Ischemic Cardiomyopathy

```{r}
knitr::include_graphics(
"enhancedvolcanoiscochemvsControl_1.0.jpeg"
)
```

## Non-Ischemic Cardiomyopathy

```{r}
knitr::include_graphics(
"enhancedvolcanoDCvsControl_1.0.jpeg"
)
```

## Boxplot Analysis

## Control vs Non-Ischemic CM

```{r}
knitr::include_graphics(
"Boxplot_Control_vs_Dilated.png"
)
```

## Control vs Ischemic CM

```{r}
knitr::include_graphics(
"Boxplot_Control_vs_Ischemic.png"
)
```

# Biological Interpretation

Several cardiomyopathy-associated genes exhibited differential expression patterns.

Key genes identified included:

-   TTN
-   ACTN2
-   DES
-   FLNC
-   CRYAB
-   FHOD3
-   PDLIM3
-   PDLIM5
-   MYOZ2

These genes are involved in:

-   Sarcomere organization
-   Cytoskeletal integrity
-   Cardiac contractility
-   Mechanical signaling
-   Heart failure progression

# Discussion

Integration of six independent transcriptomic studies substantially improved statistical power and biological consistency.

ComBat correction reduced platform-specific batch effects while Limma identified robust disease-associated transcriptional signatures.

PCA and heatmap analyses demonstrated clear separation between disease and control samples.

# Conclusion

This study successfully integrated six GEO microarray datasets to identify transcriptional alterations associated with cardiomyopathy.

Major analytical steps included:

1.  GEO Data Collection
2.  Gene Annotation
3.  Dataset Integration
4.  Batch Effect Correction
5.  Differential Expression Analysis
6.  PCA
7.  Heatmap Analysis
8.  Volcano Plot Analysis

The identified genes represent potential biomarkers and therapeutic targets for future experimental validation.

# Reproducibility

Source scripts used in this project:

-   download_data.R
-   cardiomyopathy_analysis.R

# Session Information

```{r}
sessionInfo()
```

\`\`\`
